Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-31 analysis/paper/paper.qmd
Changed numbers to codes

# create column for "n = xx "
tcsa_raw <- tcsa %>% 
  group_by(raw_materials_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
  dplyr::select(min_tcsa,
         raw_materials_top, 
       label_raw,
       n) %>% 
  distinct_all()

#
mydata_ages_na <- mydata_ages %>%
  drop_na(age)

@@

# code for numbers in the text
total_site <- nrow(korean_archaeological_site_locations)
total_sp <- nrow(mydata)
total_ass <- nrow(mydata_ages)
n_sites <- length(unique(tcsa$Site_name))
min_n_points <- nrow(tcsa_site) 
# should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
# n_radiocarbon_ages <- NULL # we only have one age per assemblage
n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
n_omit <- nrow(tcsa_omit)
n_ass <- length(unique(tcsa$Assemblage))
n_raw <- nrow(tcsa_raw)
```

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values for each raw material type to examine the interaction of raw material types and TCSA values.

`r n_raw`

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.

@@

We used radiocarbon ages to explore the temporal pattern of the function of stemmed points. We excluded assemblages that have no radiocarbon dates. We arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.

@@

To explore spatial patterns in TCSA values, we retained only sites that contains more than four stemmed points. This resulted in a sample of `r n_sites` sites and `r min_n_points` points. We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites. I THINK WE NEED SOMETHING MORE IN HERE. LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.

@@

Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks. This may be due to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times \[CITE PAPERS ON ARCHAEOLOGY AND PATCH CHOICE MODEL\]. We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.

@@

# remove NAs
tcsa_mass_no_na <-
  tcsa %>%
 drop_na(`Weight(g)`)

>>>>

# raw material : create column for "n = xx "
tcsa_raw <- tcsa %>% 
  group_by(raw_materials_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
  dplyr::select(min_tcsa,
         raw_materials_top, 
       label_raw,
       n) %>% 
  distinct_all()

# remove NAs for weight
tcsa_mass_no_na <-
  tcsa %>%
 drop_na(`Weight(g)`)

# drop na for dates
mydata_ages_na <- mydata_ages %>%
  drop_na(age)

@@

# code for numbers in the text
total_site <- nrow(korean_archaeological_site_locations)
total_sp <- nrow(mydata)
total_ass <- nrow(mydata_ages)
n_sites <- length(unique(tcsa$Site_name))
min_n_points <- nrow(tcsa_site) 
# n_radiocarbon_ages <- NULL # we only have one age per assemblage
n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
n_omit <- nrow(tcsa_omit)
n_ass <- length(unique(tcsa$Assemblage))
n_raw <- nrow(tcsa_raw)
n_size <- nrow(tcsa_mass_no_na)
n_spatial <-nrow(tcsa_site_num) # got rid of "Other" in the dataframe
n_spatial_ass <- length(unique(tcsa_site_num$Assemblage))
```

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values.

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

@@

We used radiocarbon ages to explore the temporal pattern of the function of stemmed points. After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. The we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.

@@

We sat sites that contains fewer than 5 stemmed points aside and named them as "Other". This resulted in a sample of `r n_spatial_ass` assemblages and `r n_spatial_ass` points. We summarized the distribution of TCSA values for each site to examine spatial patterns in TCSA values. We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones. We used QGIS 3.28.0 'Firenze' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.

@@

Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks. This may be due to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging]. We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.

@@
